---
title: "GEO881_Geistlich_Cyril"
author: "Cyril Geistlich"
date: "2023-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library("zoo")
library("ggplot2")
library("ggfortify")
library("simts")
library("dplyr")
library("readr")
library("sf")
library("tmap") # to map raster data/background map
library("igraph")
library("tidygraph")
library("ggraph")
```

# Data Preparation

```{r read csv data}
# Time-Series
passengers <- read_delim("data/time-series/AirPassengers.csv", ",")
pedestrians <- read_delim("data/time-series/zurich_ped_count_daily_U15G3063864_2020.csv", ",")

# Network-Analysis
csv_street_ny <- read_delim("data/network-analysis/newyork_street.csv", ",")
csv_street_zh <- read_delim("data/network-analysis/zurich_street.csv", ",")

```
```{r read shp data}
# Read Zurich SHP
shp_node_zh <- read_sf("data/network-analysis/shapefile/zurich/zurich_node.shp")
shp_street_zh <- read_sf("data/network-analysis/shapefile/zurich/zurich_street.shp")

# Read New York SHP
shp_node_ny <- read_sf("data/network-analysis/shapefile/newyork/newyork_nodes.shp")
shp_street_ny <- read_sf("data/network-analysis/shapefile/newyork/newyork_street.shp")
```


# Task 1: Explore the Data
## 1.1 Passenger Data
```{r prep passenger }
#Rename Columnsreate summary, add Id for visualisation with trend
passengers <- passengers |>
  rename(passengers = `#Passengers`,
         date = Month) |>
  mutate(id = row_number()) |>
  mutate(date = as.Date(paste(date, "01", sep = "-")))

```

```{r Explore passengers}
summary(passengers)

#Visualize Data
ggplot(passengers,aes(x = date, y = passengers)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 4))
```
## 1.2 Pedestrians
```{r prep pedestrians}
#Rename Columnsreate summary, add Id for visualisation with trend
pedestrians <- pedestrians |>
  rename(date = DATE,
         total = PED_TOTAL) |>
  mutate(id = row_number())
```

```{r Explore pedestrians}
summary(pedestrians)

#Visualize Data
ggplot(pedestrians,aes(x = date, y = total)) + 
  geom_point() +
  geom_smooth(method = "gam", col = "red") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10))
```
## 1.3 Shapefiles
### 1.3.1 Read Shapefiles

```{r prep shp}
shp_node_ny <- shp_node_ny |>
  rename(street_count = street_cou) |>
  mutate(id = row_number())

shp_street_ny <- shp_street_ny |>
  rename(from = from_,) |>
  mutate(id = row_number())

shp_node_zh |>
  rename(street_count = street_cou) |>
  mutate(id = row_number())
  

shp_street_zh <- shp_street_zh |>
  rename(from = from_,) |>
  mutate(id = row_number())
  

```

### 1.3.2 Remove Double streets

```{r remove doubles}
shp_street_ny <- shp_street_ny |>
  group_by(from,to) |>
  filter(length == min(length)) |>
  ungroup()

shp_street_zh <- shp_street_zh |>
  group_by(from,to) |>
  filter(length == min(length)) |>
  ungroup()

csv_street_ny <- csv_street_ny |>
  group_by(s,e) |>
  filter(seg_length == min(seg_length)) |>
  ungroup()

csv_street_zh <- csv_street_zh |>
  group_by(s,e) |>
  filter(seg_length == min(seg_length)) |>
  ungroup()
```

## 1.4 Visualise SHP
### 1.4.1 New York

```{r NY SHP}
#Set tmap_mode to "view" to get an interactive map
tmap_mode("plot")

tm_shape(shp_node_ny) +
  tm_dots() +
  tm_shape(shp_street_ny) +
  tm_lines(col ="highway") +
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")
```

### 1.4.2 Zurich Shapefiles

```{r Zurich SHP}
tm_shape(shp_node_zh) +
  tm_dots() +
  tm_shape(shp_street_zh) +
  tm_lines(col ="highway") +
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")
```


# Task 2: Network Analysis (50 points)

## 2.1 Create Street Grapshg and dual Graphs (5 points)

### 2.1.1 Undirected Unweighted Street Graph



```{r undirected unweighted street graph zh}
edges_zh <- csv_street_zh |>
  select(s, e, seg_length) |>
  as_tibble()

graph_zh <- tbl_graph(edges_zh, directed = F)

nodes_zh <- union(edges_zh$s,edges_zh$e)

edges_zh |> ggraph(layout = 'kk') +
  geom_edge_link() +
  geom_node_point(size = 1, colour = "steelblue") +
  theme_graph()
  
```

```{r undirected unweighted street graph ny}
edges_ny <- csv_street_ny |>
  select(s, e, seg_length) |>
  as_tibble()

graph_ny <- tbl_graph(edges_ny, directed = F)

nodes_ny <- union(edges_ny$s,edges_ny$e)

graph <- edges_ny |> ggraph(layout = 'kk') +
  geom_edge_link() +
  geom_node_point(size = 1, colour = "steelblue") +
  theme_graph()
```
Note the outliers on the right hand side with no connection to the graph

## 2.1.2 Undirected Weighted Street Graph

```{r Undirected Weighted Street Graph NY}
edges_ny |>
  ggraph(layout = 'kk') +
  geom_edge_link(aes(width = edges_ny$seg_length)) +
  geom_node_point(size = 1, colour = "steelblue") +
  geom_node_text(aes(label = sort(nodes_ny)), colour = "white", vjust = 0.4, size = 0) +
  theme_graph()
```

```{r Undirected Weighted Street Graph ZH}
edges_zh |>
  ggraph(layout = 'kk') +
  geom_edge_link(aes(width = edges_zh$seg_length)) +
  geom_node_point(size = 1, colour = "steelblue") +
  geom_node_text(aes(label = sort(nodes_zh)), colour = "white", vjust = 0.4, size = 0) +
  theme_graph()
```

## 2.2 Centrality Calculation of the Graphs (15 points)
### 2.2.1 Degree

```{r degree zh unweighted}
edges_zh |>
  mutate(centrality = centrality_degree(weights = NULL, mode = "in", normalized=FALSE)) |>
  ggraph(layout = 'kk') +
  geom_edge_link() +
  geom_node_point(size = 1, colour = "steelblue") +
  theme_graph()
```
```{r degree zh weighted}

edges_zh |>
  mutate(centrality = centrality_degree()) |>
  ggraph(layout = 'kk') +
  geom_edge_link(aes(width = edges_zh$seg_length)) +
  geom_node_point(size = 1, colour = "steelblue") +
  geom_node_text(aes(label = sort(nodes_zh)), colour = "white", vjust = 0.4, size = 0) +
  theme_graph()
```

```{r degree NY}
edges_ny |>
  ggraph(layout = 'kk') +
  geom_edge_link(aes(width = edges_ny$seg_length)) +
  geom_node_point(size = 1, colour = "steelblue") +
  geom_node_text(aes(label = sort(nodes_ny)), colour = "white", vjust = 0.4, size = 0) +
  theme_graph()
```
### 2.2.2 Closeness
```{r closeness NY}
edges_ny |>
  ggraph(layout = 'kk') +
  geom_edge_link(aes(width = edges_ny$seg_length)) +
  geom_node_point(size = 1, colour = "steelblue") +
  geom_node_text(aes(label = sort(nodes_ny)), colour = "white", vjust = 0.4, size = 0) +
  theme_graph()
```
### 2.2.3 Betweenness




## 2.3 Visual and Numerical Analytics (30 points)

# Task 3: Timer Series Analysis (40 points)
## 3.1 Analytics of Air Passenger Data (20 points)
## 3.2 Analytics of the 2020 Data (20 points)



