---
title: "GEO881_Geistlich_Cyril"
author: "Cyril Geistlich"
date: "2023-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library("zoo")
library("ggplot2")
library("ggfortify")
library("simts")
library("dplyr")
library("readr")
library("sf")
library("tmap") # to map raster data/background map
library("igraph")
library("tidygraph")
library("ggraph")
library(forecast)

```

# Data Preparation

```{r read csv data}
# Time-Series
passengers <- read_delim("data/time-series/AirPassengers.csv", ",")
pedestrians <- read_delim("data/time-series/zurich_ped_count_daily_U15G3063864_2020.csv", ",")

# Network-Analysis
csv_street_ny <- read_delim("data/network-analysis/newyork_street.csv", ",")
csv_street_zh <- read_delim("data/network-analysis/zurich_street.csv", ",")

```
```{r read shp data}
# Read Zurich SHP
shp_node_zh <- read_sf("data/network-analysis/shapefile/zurich/zurich_node.shp")
shp_street_zh <- read_sf("data/network-analysis/shapefile/zurich/zurich_street.shp")

# Read New York SHP
shp_node_ny <- read_sf("data/network-analysis/shapefile/newyork/newyork_nodes.shp")
shp_street_ny <- read_sf("data/network-analysis/shapefile/newyork/newyork_street.shp")
```

# Task 1: Explore the Data
## 1.1 Passenger Data
```{r prep passenger }
#Rename Columnsreate summary, add Id for visualisation with trend
passengers <- passengers |>
  rename(passengers = `#Passengers`,
         date = Month) |>
  mutate(id = row_number()) |>
  mutate(date = as.Date(paste(date, "01", sep = "-")))

```


## 1.2 Pedestrians
```{r prep pedestrians}
#Rename Columnsreate summary, add Id for visualisation with trend
pedestrians <- pedestrians |>
  rename(date = DATE,
         total = PED_TOTAL) |>
  mutate(id = row_number())
```


## 1.3 Shapefiles
### 1.3.1 Read Shapefiles

```{r prep shp}
shp_node_ny <- shp_node_ny |>
  rename(street_count = street_cou)

shp_street_ny <- shp_street_ny |>
  rename(from = from_,) |>
  mutate(id = row_number())

shp_node_zh <- shp_node_zh |>
  rename(street_count = street_cou) 
  
shp_street_zh <- shp_street_zh |>
  rename(from = from_,) |>
  mutate(id = row_number())
  
```

### 1.3.2 Remove Double streets

```{r remove doubles}
shp_street_ny <- shp_street_ny |>
  group_by(from,to) |>
  filter(length == min(length)) |>
  ungroup()

shp_street_zh <- shp_street_zh |>
  group_by(from,to) |>
  filter(length == min(length)) |>
  ungroup()

csv_street_ny <- csv_street_ny |>
  group_by(s,e) |>
  filter(seg_length == min(seg_length)) |>
  ungroup()

csv_street_zh <- csv_street_zh |>
  group_by(s,e) |>
  filter(seg_length == min(seg_length)) |>
  ungroup()
```

## 1.4 Visualise SHP
### 1.4.1 New York

```{r}
#Set tmap_mode to "view" to get an interactive map
tmap_mode("plot")
```


```{r NY SHP}
tm_shape(shp_node_ny) +
  tm_dots() +
  tm_shape(shp_street_ny) +
  tm_lines(col ="highway") +
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")
```

### 1.4.2 Zurich Shapefiles

```{r Zurich SHP}
tm_shape(shp_node_zh) +
  tm_dots() +
  tm_shape(shp_street_zh) +
  tm_lines(col ="highway") +
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")
```


# Task 2: Network Analysis (50 points)
## 2.1 Create Street Graphs and dual Graphs (5 points)
## 2.2 Centrality (degree, closeness, betweenness) Calculation of the Graphs (15 points)
## 2.3 Visual and Numerical Analytics (30 points)

```{r preparation}
# Use start end combination as id
csv_street_ny <- csv_street_ny |> mutate(s_e = paste0(pmin(s, e), pmax(s, e)))
csv_street_zh <- csv_street_zh |> mutate(s_e = paste0(pmin(s, e), pmax(s, e)))

ny <- csv_street_ny |>
  arrange(seg_length) |>
  distinct(s_e, .keep_all=TRUE)

zh <- csv_street_zh |>
  arrange(seg_length) |>
  distinct(s_e, .keep_all=TRUE)

```

### 2.1.1 Undirected Unweighted Street Graph

```{r undirected unweighted street graphs}
# edge information
ny_edge <- data.frame(from=ny$s, to=ny$e, weight=ny$seg_length)
zh_edge <- data.frame(from=zh$s, to=zh$e, weight=zh$seg_length)

# node information
ny_node <- union(ny$s, ny$e)
zh_node <- union(zh$s, zh$e)

# Creating the street graphs
ny_street <- graph_from_data_frame(vertices=ny_node, d=ny_edge, directed=FALSE) |> as_tbl_graph()
zh_street <- graph_from_data_frame(vertices=zh_node, d=zh_edge, directed=FALSE) |> as_tbl_graph()

```

```{r}
zh_street_unweighted <- zh_street |>
  mutate(degree = centrality_degree(weights=NULL),           # degree centrality
         closeness = centrality_closeness(weights=NULL),     # closeness centrality
         betweenness = centrality_betweenness(weights=NULL)) # betweenness centrality

ny_street_unweighted <- ny_street |>
  mutate(degree = centrality_degree(weights=NULL),           # degree centrality
         closeness = centrality_closeness(weights=NULL),     # closeness centrality
         betweenness = centrality_betweenness(weights=NULL)) # betweenness centrality
```

```{r}
zh_nodes_street_unweighted <- zh_street_unweighted |> 
  activate(nodes) |> 
  data.frame() |> 
  mutate(name=as.numeric(name))

ny_nodes_street_unweighted <- ny_street_unweighted |> 
  activate(nodes) |> 
  data.frame() |> 
  mutate(name=as.numeric(name))

```

```{r}
sf_nodes_zh_street_unweighted <- shp_node_zh |> 
  left_join(zh_nodes_street_unweighted, by=c("osmid"="name"))

sf_nodes_ny_street_unweighted <- shp_node_ny |> 
  left_join(ny_nodes_street_unweighted, by=c("osmid"="name"))
```

```{r}
zh_plot_street_unweighted <- ggplot() + 
  geom_sf(data=shp_street_zh, size=0.25) +
  geom_sf(data=sf_nodes_zh_street_unweighted, aes(col=as.factor(degree)), size=0.5) +
  scale_color_brewer(palette="YlOrRd") +
  labs(title="Zurich: Unweighted street graph", col="degree centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_street_unweighted <- ggplot() + 
  geom_sf(data=shp_street_ny, size=0.25) +
  geom_sf(data=sf_nodes_ny_street_unweighted, aes(col=as.factor(degree)), size=0.5) +
  scale_color_brewer(palette="YlOrRd") +
  labs(title="New York: Unweighted street graph", col="degree centrality") +
  theme_dark()

zh_plot_street_unweighted
ny_plot_street_unweighted 
```

```{r}

zh_plot_street_unweighted <- ggplot() + 
  geom_sf(data=shp_street_zh, size=0.25) +
  geom_sf(data=sf_nodes_zh_street_unweighted, aes(col=closeness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 0.00004) +
  labs(title="Zurich: Unweighted street graph", col="closeness centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_street_unweighted <- ggplot() + 
  geom_sf(data=shp_street_ny, size=0.25) +
  geom_sf(data=sf_nodes_ny_street_unweighted, aes(col=closeness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 0.2) +
  labs(title="New York: Unweighted street graph", col="closeness centrality") +
  theme_dark()


zh_plot_street_unweighted
ny_plot_street_unweighted 
```

```{r}
zh_plot_street_unweighted <- ggplot() + 
  geom_sf(data=shp_street_zh, size=0.25) +
  geom_sf(data=sf_nodes_zh_street_unweighted, aes(col=betweenness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 100000) +
  labs(title="Zurich: Unweighted street graph", col="betweenness centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_street_unweighted <- ggplot() + 
  geom_sf(data=shp_street_ny, size=0.25) +
  geom_sf(data=sf_nodes_ny_street_unweighted, aes(col=betweenness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 400000) +
  labs(title="New York: Unweighted street graph", col="betweenness centrality") +
  theme_dark()


zh_plot_street_unweighted
ny_plot_street_unweighted 
```

# 2.1.2 Dual Graph

```{r}
create_dual_graph <- function(data) {
  node <- data$FID
  edges <- data.frame(from = numeric(), to = numeric())

  # Test all posibilites. 
  for (FID1 in 1:length(node)) {
    for (FID2 in FID1:length(node)) {
      if (FID1 != FID2) { # Dont match to identical points
        FID1_s <- data$s[FID1]
        FID1_e <- data$e[FID1]
        FID2_s <- data$s[FID2]
        FID2_e <- data$e[FID2]

        if ((FID1_s == FID2_s) || (FID1_s == FID2_e) || (FID1_e == FID2_s) || (FID1_e == FID2_e)) {
          edges <- rbind(edges, data.frame(from = data$FID[FID1], to = data$FID[FID2]))
        }}}}

  edges <- edges |>
    mutate(s_e = paste0(pmin(from, to), pmax(from, to))) |>
    distinct(s_e, .keep_all = TRUE) |>
    select(from, to)

  graph <- graph_from_data_frame(d = edges, vertices = node, directed = FALSE) |>
    as_tbl_graph()

  return(graph)
}

```

```{r}
ny_dual <- create_dual_graph(ny)
zh_dual <- create_dual_graph(zh)
```

```{r}
zh_dual <- zh_dual |>
  mutate(degree = centrality_degree(weights=NULL),           # degree centrality
         closeness = centrality_closeness(weights=NULL),     # closeness centrality
         betweenness = centrality_betweenness(weights=NULL)) # betweenness centrality

ny_dual <- ny_dual |>
  mutate(degree = centrality_degree(weights=NULL),           # degree centrality
         closeness = centrality_closeness(weights=NULL),     # closeness centrality
         betweenness = centrality_betweenness(weights=NULL)) # betweenness centrality
```

```{r}
# dual graphs
zh_nodes_dual <- zh_dual |> 
  activate(nodes) |> 
  data.frame() |> 
  mutate(name=as.numeric(name))

ny_nodes_dual <- ny_dual |> 
  activate(nodes) |> 
  data.frame() |> 
  mutate(name=as.numeric(name))
```

```{r}
sf_streets_zh_dual <- shp_street_zh |>
  mutate(FID = id-1) |>
  left_join(zh_nodes_dual, by=c("FID"="name"))

sf_streets_ny_dual <- shp_street_ny |>
  mutate(FID = id-1) |>
  left_join(ny_nodes_dual, by=c("FID"="name"))
```


```{r}
zh_plot_dual <- ggplot() + 
  geom_sf(data=sf_streets_zh_dual, aes(col=as.factor(degree))) +
  scale_color_brewer(palette="YlOrRd") +
  labs(title="Zurich: Dual graph", col="degree centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_dual <- ggplot() + 
  geom_sf(data=sf_streets_ny_dual, aes(col=as.factor(degree))) +
  scale_color_brewer(palette="YlOrRd") +
  labs(title="New York: Dual graph", col="degree centrality") +
  theme_dark()

zh_plot_dual
ny_plot_dual
```

```{r}
zh_plot_dual <- ggplot() + 
  geom_sf(data=sf_streets_zh_dual, aes(col=closeness)) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 0.00004) +
  labs(title="Zurich: Dual graph", col="closeness centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_dual <- ggplot() + 
  geom_sf(data=sf_streets_ny_dual, aes(col=closeness)) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 0.3) +
  labs(title="New York: Dual graph", col="closeness centrality") +
  theme_dark()

zh_plot_dual
ny_plot_dual 
```

```{r}

zh_plot_dual <- ggplot() + 
  geom_sf(data=sf_streets_zh_dual, aes(col=betweenness)) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 300000) +
  labs(title="Zurich: Dual graph", col="betweenness centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_dual <- ggplot() + 
  geom_sf(data=sf_streets_ny_dual, aes(col=betweenness)) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 1000000) +
  labs(title="New York: Dual graph", col="betweenness centrality") +
  theme_dark()
  

zh_plot_dual
ny_plot_dual 
```

## 2.1.3 Undirected Weighted Street Graph

```{r}
zh_street_weighted <- zh_street |>
  mutate(degree = centrality_degree(weights=weight),           # degree centrality
         closeness = centrality_closeness(weights=weight),     # closeness centrality
         betweenness = centrality_betweenness(weights=weight)) # betweenness centrality

ny_street_weighted <- ny_street |>
  mutate(degree = centrality_degree(weights=weight),           # degree centrality
         closeness = centrality_closeness(weights=weight),     # closeness centrality
         betweenness = centrality_betweenness(weights=weight)) # betweenness centrality
```

```{r}
# weighted street graphs
zh_nodes_street_weighted <- zh_street_weighted |> 
  activate(nodes) |> 
  data.frame() |> 
  mutate(name=as.numeric(name))

ny_nodes_street_weighted <- ny_street_weighted |> 
  activate(nodes) |> 
  data.frame() |> 
  mutate(name=as.numeric(name))
```

```{r}
sf_nodes_zh_street_weighted <- shp_node_zh |>
  left_join(zh_nodes_street_weighted, by=c("osmid"="name"))

sf_nodes_ny_street_weighted <- shp_node_ny |>
  left_join(ny_nodes_street_weighted, by=c("osmid"="name"))
```

```{r}

zh_plot_street_weighted <- ggplot() + 
  geom_sf(data=shp_street_zh, size=0.25) +
  geom_sf(data=sf_nodes_zh_street_weighted, aes(col=degree), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 175) +
  labs(title="Zurich: Weighted street graph", col="degree centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_street_weighted <- ggplot() + 
  geom_sf(data=shp_street_ny, size=0.25) +
  geom_sf(data=sf_nodes_ny_street_weighted, aes(col=degree), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 1000) +
  labs(title="New York: Weighted street graph", col="degree centrality") +
  theme_dark()


zh_plot_street_weighted
ny_plot_street_weighted  
```

```{r}

zh_plot_street_weighted <- ggplot() + 
  geom_sf(data=shp_street_zh, size=0.25) +
  geom_sf(data=sf_nodes_zh_street_weighted, aes(col=closeness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 0.000001) +
  labs(title="Zurich: Weighted street graph", col="Closeness centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_street_weighted <- ggplot() + 
  geom_sf(data=shp_street_ny, size=0.25) +
  geom_sf(data=sf_nodes_ny_street_weighted, aes(col=closeness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 0.010) +
  labs(title="New York: Weighted street graph", col="Closeness centrality") +
  theme_dark()


zh_plot_street_weighted
ny_plot_street_weighted
```


```{r}
zh_plot_street_weighted <- ggplot() + 
  geom_sf(data=shp_street_zh, size=0.25) +
  geom_sf(data=sf_nodes_zh_street_weighted, aes(col=betweenness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 80000) +
  labs(title="Zurich: Weighted street graph", col="betweenness centrality") +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ny_plot_street_weighted <- ggplot() + 
  geom_sf(data=shp_street_ny, size=0.25) +
  geom_sf(data=sf_nodes_ny_street_weighted, aes(col=betweenness), size=0.5) +
  scale_color_gradient2(low = "yellow", mid = "red", high = "purple", midpoint = 500000) +
  labs(title="New York: Weighted street graph", col="betweenness centrality") +
  theme_dark()


zh_plot_street_weighted
ny_plot_street_weighted
```

# Task 3: Timer Series Analysis (40 points)
## 3.1 Analytics of Air Passenger Data (20 points)

```{r Explore passengers}
summary(passengers)

#Visualize Data
ggplot(passengers,aes(x = date, y = passengers)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12))
```

```{r}
# Create TimeSeries Object
(passengers_ts <- ts(passengers$passengers, frequency = 12, start = min(year(passengers$date))))
```

```{r}
# Convert the month number to month name
month_names <- month.abb[cycle(passengers_ts)]

# Create an ordered factor for month names with correct ordering
month_names <- factor(month_names, levels = month.abb)

# Create a boxplot of passenger values grouped by month
boxplot(passengers_ts ~ month_names,
        xlab = "Month",
        ylab = "passengers",
        main = "Monthly Values")
```

```{r}
# Apply ACF to the time series
acf_passengers <- acf(passengers_ts, lag.max = 12)

# Plot the ACF
plot(acf_passengers, main = "Autocorrelation Function (ACF) of Passenger Time Series")
```
```{r}
# Apply STL decomposition to the time series
stl_passengers_add <- stl(passengers_ts, s.window = "periodic")
stl_passengers_mult <- stl(log(passengers_ts)- min(passengers_ts) + 1, s.window = "periodic")

# Plot the decomposed components
plot(stl_passengers_add, main = "Additive decomposition")

# Access the components: trend, seasonal, and remainder
trend_add <- stl_passengers_add$time.series[, "trend"]
seasonal_add <- stl_passengers_add$time.series[, "seasonal"]
remainder_add <- stl_passengers_add$time.series[, "remainder"]
```
```{r}
# Plot the decomposed components
plot(stl_passengers_mult,main = "Multiplicative decomposition")

# Access the components: trend, seasonal, and remainder
trend_mult <- stl_passengers_mult$time.series[, "trend"]
seasonal_mult <- stl_passengers_mult$time.series[, "seasonal"]
remainder_mult <- stl_passengers_mult$time.series[, "remainder"]
```
```{r}
par(mfrow = c(1,2))
plot(stl_passengers_add$time.series[,3], 
     col = "blue", main = 'Remainder', ylab = "")

qqnorm(remainder_add, main="Q-Q plot additive model")
qqline(remainder_add, col="red")
```


```{r}
par(mfrow = c(1,2))
plot(stl_passengers_mult$time.series[,3], 
     col = "blue", main = 'Remainder', ylab = "")
qqnorm(remainder_mult, main="Q-Q plot multiplicative model")
qqline(remainder_mult, col="red")
```

```{r}
# Apply exponential smoothing
ets_model <- ets(passengers_ts)

summary(ets_model)
```
```{r}
autoplot(ets_model)
```
```{r}
(forecast_ets_model <- forecast(ets_model) |> autoplot())
```

```{r}
arima_model <- auto.arima(passengers_ts)

autoplot(arima_model$residuals)
```

```{r}
# Generate forecasts
forecast_arima <- forecast(arima_model, h = 24)  # Adjust 'h' to the desired number of future time points

# Plot the forecasted values
plot(forecast_arima, main = "ARIMA Forecast")

# Print the forecasted values
print(forecast_arima)
```


## 3.3.2 Analytics of the 2020 Data (20 points)

```{r Explore pedestrians}
summary(pedestrians)

#Visualize Data
ggplot(pedestrians,aes(x = date, y = total)) + 
  geom_point() +
  geom_smooth(method = "gam", col = "red") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10))
```

Judging from this plot I assume 1 break in the pedestrians data set. 
The break occurs around April 2020 when first lock down measures where implemented. 

```{r}
library(changepoint)
library(strucchange)

```

```{r}
# Create Timeseries object
pedestrian_ts <- ts(pedestrians$total, frequency = 365, start = min(date(pedestrians$date)))
```

```{r}
#changes in mean, variance & mean variance
cp_mean <- cpt.mean(pedestrian_ts)
cp_var <- cpt.var(pedestrian_ts)
cp_var_mean <- cpt.meanvar(pedestrian_ts)

# Visualize
par(mfrow=c(3,1))
plot(cp_mean, main="Mean")
plot(cp_var, main="Variance")
plot(cp_var_mean, main="Variance and mean")

breakpoints <- cp_mean@cpts
print(breakpoints)

breakpoints <- cp_var@cpts
print(breakpoints)


breakpoints <- cp_var_mean@cpts
print(breakpoints)

```
with the chanepoint library and methods we can find one breakpoint at index = 14 (May 24th) 

```{r}
breakpoints <- 144

ggplot(pedestrians) +
  geom_line(aes(x = id, y = total)) +
  geom_vline(xintercept = breakpoints, color = "red") +
  ylab("Pedestrians") 
  
```


```{r}
# use strucchange to find breakpoints
strucchange <- breakpoints(pedestrian_ts ~ 1, breaks=5)
summary(strucchange)
plot(strucchange)
```
```{r}
breakpoints <- strucchange$breakpoints

ggplot(pedestrians) +
  geom_line(aes(x = id, y = total)) +
  geom_vline(xintercept = breakpoints, color = "red") +
  ylab("Pedestrians") 
  
```

With Strucchange we find 3 breakpoints. At day 75 (March 16th), day 134 (May 14th) and day 282 (September 10th).

```{r}

pedestrian_ts <- ts(pedestrians$total, frequency = 1)
breakpoints <- c(134, 365)

# Loop (initally all breakpoints were used)
for (i in 1:length(breakpoints)) {
  # Subset the data at each breakpoint
  subset_data <- pedestrians$total[1:breakpoints[i]]
  
  # Convert the subset to a time series
  subset_ts <- ts(subset_data, frequency = 1)
  
  # Perform autocorrelation test on the subset
  acf_result <- forecast::Acf(subset_ts)
  
  # Plot the autocorrelation function
  plot(acf_result, main = paste("Autocorrelation - Subset", i))
}
```

